---
title: "Exercise 18.1 - Measuring predictor importance"
output: html_document:
keep_md: true 

The “churn” data set was developed to predict telecom customer churn based on 
information about their account. It contains 20 variables, with the 19
predictors including continuous and factor variables that describe an individual
account. The response churn has two levels "yes" and "no". This script....

(a) Examining the correlation between predictors
-------------------------------------------------------------------------------
```{r}
# load packages
library(AppliedPredictiveModeling)
library(C50)
library(corrplot)
library(car)

# load data
data(churn)
str(churnTrain)

# correlation between the continuous variables the the 
numeric_vars <- sapply(churnTrain, is.numeric)
corrs <- cor(churnTrain[numeric_vars])
corrplot(corrs, method = "number", tl.cex = 0.75)
```

Finding a perfect colinearity between the of four pairs of total charge and 
total minutes variables is of course no surprise, phone charges are not random 
variables! However it is somewhat surprising that there is no linear relation 
between any other pairs of variables. A scatterplot matrix of a subset of 
the continuous variables reveals the extent of any pairwise relationship and 
how certain variables contain large amounts of zeros.

```{r, echo=FALSE}
pairs(~.,data=churnTrain[numeric_vars][1:4], main="Scatterplot Matrix")
# make conditional on churn
```

(b) Assessing the importance of categorical predictors
-------------------------------------------------------------------------------

